---
title: "Delta approx. of variance of pseudobulk statistics"
subtitle: 'Theory vs simulations'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
  self_contained: false
---



<!---

cd /Users/gabrielhoffman/Dropbox/projects/dreamlet/theory_sims


# laptop
rmarkdown::render("theory_sims.Rmd")


--->

Perform simulations to test the theoretically derived sampling variance for the mean and variance of CPM and log CPM.  Note that the number of cells observed per sample is plotted on the x-axis on a log scale, and the variances are plotted on the y-axis on a log scale.

We note that a constant prior count  of 0.05 is used for all cases here for simplicity.  This causes a small bias in the estimates of the mean values.  Dreamlet implements an adaptive prior count approach, so this is not an issue in real data.    

```{r knitr, echo=FALSE, message=FALSE}
library(ggplot2)
suppressPackageStartupMessages(library(knitr))
options(xtable.type="html")

knitr::opts_chunk$set(
  echo=TRUE,
  warning=FALSE,
  message=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  cache.lazy = FALSE,
  dev = c("png", "pdf"), 
  fig.width=7, fig.height=7)

# options(markdown.HTML.stylesheet = 'css/custom.css')
```

```{r msims, fig.width=5, fig.height=5}
library(tidyverse)
library(purrr)
library(parallel)
# lib.size = rep(2000, 10000)

lib.size = matrix(0, nrow=10000, ncol=3)
lib.size[,1] = runif(10000, 1000, 1000)
lib.size[,2] = runif(10000, 5000, 5000)
lib.size[,3] = runif(10000, 15000, 15000)

# Simulate prop
# mu = 0.001
# sigmaSq = 1e-7
mu = 0.001
sigmaSq = 2e-7
alpha = mu^2 / sigmaSq
beta = mu / sigmaSq

# mu *5000

prop = rgamma(length(lib.size), shape = alpha, rate = beta)
hist(prop, main="Distribution of expression fractions", xlim=c(0, max(prop)))

# cells
df = mclapply(c(3,5,7, seq(10, 150, by=15), seq(200, 800, by=50), 1000), function(nc){
	# replicates
	map_df(seq(1000), function(repl){
		# Donor
		map_df(seq(3), function(d){
			i = sample.int(nrow(lib.size), nc)
			x = rpois(length(i), prop[i]*lib.size[i,d])

			tibble(ncells = nc, 
				Repl = repl, 
				Donor = factor(lib.size[1,d], levels=lib.size[1,]), 
				x = x, 
				lib.size = lib.size[i,d])
		})
	})
}, mc.cores=4)
df = do.call(rbind, df)
 
# summary for each ncells and Repl
df2 = df %>%
	group_by(ncells, Repl, Donor) %>%
	summarize(Counts = sum(x) + .05, 
			mean.lib.size = mean(lib.size), 
			zeta = mean(lib.size^2),
			per.cell.cpm = mean((x+0.05) / lib.size)*1e6) %>%
	group_by(ncells, Repl, Donor) %>%
	summarize(sum = sum(Counts),
				mu.cpm = mean(Counts / (ncells*mean.lib.size) *1e6),
				mu.lcpm = mean(log(Counts / (ncells*mean.lib.size) *1e6)),
				mean.lib.size = mean.lib.size, 
				zeta = zeta, 
				per.cell.cpm = mean(per.cell.cpm)) 

# average over replicates
df3 = df2 %>% 
		group_by(ncells, Donor) %>%
		summarize( v.sum = var(sum),
			mu.cpm2 = mean(mu.cpm),
			v.cpm2 = var(mu.cpm),
			mu.lcpm2 = mean(mu.lcpm),
			v.lcpm2 = var(mu.lcpm),
			mean.lib.size = mean(mean.lib.size),
			zeta = mean(zeta),
			counts = mean(sum), 
			per.cell.cpm = mean(per.cell.cpm))
```

# Plots
Black dots indicate values estimated from simulation. Red lines indicate values predicted given the true parameter value. Green lines indicate values predicted given observed counts, but using true values for `sigmaSq`
```{r plots, cache=FALSE, fig.height=3, fig.width=8}

thm = theme_bw() +
	theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) 

ggplot(df3, aes(ncells, counts)) +
	geom_point(size=2) +
	thm +
	geom_line(aes(ncells, mu*ncells*mean.lib.size), col="red", linetype="solid", size=1.2) +
	facet_wrap(~ Donor) +
	xlab("Cells per sample") +
	ylab("mean(Counts)") +
	ggtitle("mean(Counts)") +
	scale_y_log10() + 
	scale_x_log10()

ggplot(df3, aes(ncells, v.sum)) +
	geom_point(size=2) +
	thm +
	geom_line(aes(ncells, ncells*(mu*mean.lib.size + sigmaSq*zeta)), col="red", linetype="solid", size=1.2) +
	thm +
	geom_line(aes(ncells, ncells*(counts/ncells + sigmaSq*zeta)), col="green", linetype="dashed", size=1.2) +
	facet_wrap(~ Donor) +
	ggtitle("var(Counts)") +
	ylab("var(Counts)") +
	xlab("Cells per sample") +
	scale_y_log10() + 
	scale_x_log10()

ggplot(df3, aes(ncells, mu.cpm2)) +
	geom_point(size=2) +
	thm +
	geom_hline(yintercept = mu*1e6, linetype="solid", col="red", size=1.2) +
	facet_wrap(~ Donor) +
	ggtitle("mean(CPM)") +
	ylab("mean(CPM)") +
	xlab("Cells per sample") +
	ylim(mu*0.9e6, mu*1.1e6) +
	scale_x_log10()

df3 %>%
	mutate(mu.hat = counts / (ncells*mean.lib.size)) %>%
	ggplot(aes(ncells, v.cpm2)) +
	geom_point(size=2) +
	thm +
	geom_line(aes(ncells, 1e12*(mu + sigmaSq*zeta/mean.lib.size)/(mean.lib.size*ncells)), linetype="solid", col="red", size=1.2) +
	geom_line(aes(ncells, 1e12*(mu.hat + sigmaSq*zeta/mean.lib.size)/(mean.lib.size*ncells)), linetype="dashed", col="green", size=1.2)  +
	facet_wrap(~ Donor) +
	ggtitle("var(CPM)") +
	ylab("var(CPM)") +
	xlab("Cells per sample") +
	scale_y_log10() + 
	scale_x_log10()

rng = range(c(df3$mu.lcpm2, log(mu) + log(1e6)))
ggplot(df3, aes(ncells, mu.lcpm2)) +
	geom_point(size=2) +
	thm +
	geom_hline(yintercept = log(mu) + log(1e6), linetype="solid", col="red", size=1.2)  +
	facet_wrap(~ Donor) +
	ggtitle("mean(log CPM)") +
	ylab("mean(log CPM)") +
	xlab("Cells per sample") +
	scale_y_log10(limits=rng) +
	scale_x_log10()

ggplot(df3, aes(ncells, v.lcpm2)) +
	geom_point(size=2) +
	thm +
	geom_line(aes(ncells, 1/(mu*ncells*mean.lib.size) * (1 + sigmaSq*zeta/(mu*mean.lib.size))), color="red", size=1.2, linetype="solid") +

	geom_line(aes(ncells, 1/(counts) * (1 + sigmaSq*zeta/(counts/ncells))), color="green", size=1.2, linetype="dashed") +
	facet_wrap(~ Donor) +
	ggtitle("var(log CPM)")  +
	ylab("var(log CPM)") +
	xlab("Cells per sample") +
	scale_y_log10() + 
	scale_x_log10()

ggplot(df3,aes(ncells, v.lcpm2 * counts)) +
	geom_point(size=2) +
	thm +
	geom_line(aes(ncells, (1 + sigmaSq*zeta/(mu*mean.lib.size))), color="red", size=1.2, linetype="solid") +
	geom_line(aes(ncells, (1 + sigmaSq*zeta/(counts/ncells))), color="green", size=1.2, linetype="dashed") +
	ylab("Variance inflation") +
	xlab("Cells per sample") +
	facet_wrap(~ Donor) +
	scale_x_log10() +
	scale_y_continuous(limits=c(1, NA))
```


```{r plots.vif, cache=FALSE, fig.height=3, fig.width=8}
# plot VIF as a function of lib and sigmaSq
df4 = expand.grid(sigmaSq = seq(.1,100, length.out=1000)*1e-7, 
								mean.lib.size = c(1000, 5000, 15000))
df4$zeta = df4$mean.lib.size^2
df4$vif = with(df4, (1 + sigmaSq*zeta/(mu*mean.lib.size)))

ggplot(df4,aes(sigmaSq, vif)) +
	geom_line(color="red", size=1.2, linetype="solid") +
	thm +
	facet_wrap(~ mean.lib.size)+
	# scale_y_continuous(limits=c(0, NA))  +
	scale_x_log10() +
	scale_y_log10() + 
	# geom_hline(yintercept=1, color="grey30", linetype="dashed") +
	ylab("Variance inflation factor (log10)")
```













